---
title: "Gina's fish personality"
output: pdf_document
date: "2023-05-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background information

This project evaluates the effect of BMAA (a cyanotoxin) on fathead minnow behavior. Young fish were grouped in 3 treatment conditions: control (no BMAA exposure), T5 (low BMAA exposure) and T25 (moderate BMAA exposure). Behavioral responses to an open-field test were then recorded at 8 time points across development. Ethovision was used to record behavior during the open-field test and resulted in many different variables that could represent different aspects of personality. Additionally, embryos were test once for burst activity.

Our main questions with these data are:
1. Are the behaviors recorded during the open field test reflective of underlying personality traits? To test this we will evaluate the repeatability of each behavior across the 8 time points.
2. How does BMAA exposure affect behavior in the open field test? To test this we will compare performance among treatment groups.


## Methods

We first need to determine the relationships among the variables recorded by Ethovision.

```{r variables}
etho = read.csv(url("https://raw.githubusercontent.com/ginalamka/Personality_BMAA/main/EthoData_updated.csv?token=GHSAT0AAAAAACB372QA37AYRBA76MHRCWXSZD7LT4Q"), header=T, sep=",", stringsAsFactors=F)

##How correlated are the variables?
#Create dataframe with just numeric/integer values
etho.corr = etho[,-c(1:7)]
etho.corr$MeanTimeZ2 = as.numeric(etho.corr$MeanTimeZ2)
etho.corr$LatencyZ2 = as.numeric(etho.corr$LatencyZ2)
etho.corr=etho.corr[-which(is.na(etho.corr$MeanTimeZ2)),] #There are some missing data that may need to be changed to ceiling values? For now, we will ignore those.


library(Hmisc)
library(corrplot)
#Dataframe should be a matrix. 
cor = rcorr(as.matrix(etho.corr))
corrplot(cor$r,type = "upper",p.mat=cor$P, sig.level=0.05, insig = "blank",diag=F) #lots of variables correlated to differing degrees




##Do correlated variables load on PCs to suggest personality axes?
#calc principal components
pca <- prcomp(etho.corr, scale=TRUE)

#need to reverse the signs
pca$rotation <- -1*pca$rotation

#calc total variance explained by each princ component
pca$sdev^2/sum(pca$sdev^2) # first 5 PCs account for 72ish % of variance in data

#display principal components
loadings = data.frame(pca$rotation[1:25,1:5]) #all 25 variables, first 5 PCs
loadings = loadings[order(-abs(loadings$PC1), -abs(loadings$PC2), -abs(loadings$PC3), -abs(loadings$PC4), -abs(loadings$PC5)),]

#create biplot to visualize
biplot(pca$x[, 1:2], pca$rotation[, 1:2], cex=1)

```

## Interpreting the loadings (From http://strata.uga.edu/8370/lecturenotes/principalComponents.html): 
Because the sum of the squares of all loadings for an individual principal component must sum to one, we can calculate what the loadings would be if all variables contributed equally to that principal component. Any variable that has a larger loading than this value contributes more than one variable’s worth of information and would be regarded as an important contributor to that principal component.

We have 25 variables, the square root of the value if each variable contributed equally (1/25) is 0.2. So, variable loadings that are greater than 0.2 contribute a larger than expected amount to the PC.

# PC1 - fast vs slow-exploring
The following variables have large, negative loadings: MeanVel, TotDist, FreqTransZ2.Z1, Mean.Activ, Tot.Act, FreqTransZ1.Z2, Mean.Mobility, FreqZ2.
The following variables have large, positive loadings (and so have an opposite effect of PC1): VarTurnAngle, LatencyZ2, LatencyZ2360.

# PC2
Large, negative loadings: CumDur.Z1, CumDurZ1
Large, positive loadings: CumDur.Z2, CumDurZ2, MeanTimeZ2, MeanTimeZ2Zero, FreqZ2


```{r variables by age}
#### Do PCAs for each age
#keep track of variance and loadings for each PC at each age with data frame
#Age categories: 21  49  77  14 105  133 161 189
etho$Age[which(etho$Age == 15)]<-14

for(i in unique(etho$Age)){
  temp = etho[etho$Age==i,]
  temp$MeanTimeZ2 = as.numeric(temp$MeanTimeZ2)
  temp$LatencyZ2 = as.numeric(temp$LatencyZ2)
  temp=temp[-which(is.na(temp$MeanTimeZ2)),]
  pca_i <- prcomp(temp[,8:32], scale=TRUE)
  pca_i$rotation <- -1*pca_i$rotation
  var = data.frame(pca_i$sdev^2/sum(pca_i$sdev^2))
  rot = data.frame(pca_i$rotation[1:25,1])
  colnames(var) = i
  colnames(rot) = i
  if(i==21){
    vars = var
    rots = rot
  } else{
    vars = cbind(vars, var)
    rots = cbind(rots, rot)
  }
}
#Range of variance accounted for by PC1 is 25-31%
#Range of PC2 is 16-25%
vars$PC = row.names(vars)
scree.plot = gather(vars, age, value, "21":"189",factor_key =F)
ggplot(scree.plot, aes(x = as.numeric(PC), y = (value)*100, color = as.factor(age))) + 
  theme_bw() +
  geom_line() +
  geom_hline(yintercept=4) #the amount of variance each variable would contribute if all contributed the same amount: 1/25*100


###do PC loadings stay the same across ages? Spaghetti plot:
library(tidyr)
library(ggplot2)
rots$variables = row.names(rots)
plot.data = gather(rots, age, value, "21":"189", factor_key = F)
ggplot(plot.data, aes(x = as.numeric(age), y = abs(value), color = variables)) +
  geom_line() +
  theme_bw()

```

## PC conclusions
The PC loadings are too diffuse and inconsistent across time points. We will move forward with candidate variables for 3 personality traits, chosen based on the correlation matrix.



# Variables:
1. Exploration = Angular Velocity. The relative turn angle / time difference ; degrees per second; can be + or -. Measures the speed of change in direction of movement - used to assess turn bias or circular tendency and abnormalities of behavior. NOTE - turn angle is sensitive to small movements of body points. if distance moved is very small, turn angle can get high, unrealistic values (additionally, consecutive turns can have high values). To combat this, let's scale distance moved

2. Activity = Mobility. The percentage of pixel change between current sample and previous sample *detected in the subject only* ; ranges from 0 - 100%. NOTE - mobility depends on size of subject only, not on arena size. small fish results in small number of pixels that change, therefore small movements results in high change

3. Boldness = Cumulative duration in the center.

```{r turning angle rpt}
etho = read.csv("/Users/kelseymccune/Documents/GitHub/Personality_BMAA/EthoData_updated.csv", header=T, sep=",", stringsAsFactors=F)

#test fit of Gaussian distribution
library(ggpubr)
ggqqplot(log(abs(scale(etho$MeanAngVel, center = T, scale = T))))
shapiro.test(log(abs(scale(etho$MeanAngVel, center = T, scale = T))))

library(lme4)
ta1 = lmer(log(abs(scale(MeanAngVel, center = T, scale = T))) ~ Treatment + Age + (1|FishName) + (1|Clutch) + (1|Paternity), data = etho)
summary(ta1) #fit is singular because clutch random effect accounts for 0 variance
ggqqplot(ta1@u)
plot(ta1)

# Maybe MeanTurnAngle will fit better? Why do values range from -16 to +16? Description of this variable on pg. 63 of Manual pdf
# Do we have absolute or relative mean turn angle/angular velocity?
ta2 = lmer(log(abs(MeanTurnAngle)) ~ Treatment + Age + (1|FishName) + (1|Clutch) + (1|Paternity), data = etho)
summary(ta2)
ggqqplot(ta2@u)
plot(ta2)

```


## Rethinking exploration

The data for turning angle are summarized from Relative turning angle (where the sign of the turn is taken in to account). This means that mean of turning angle within a trial are not reflective of the number and degree of turns individuals make because turns in different directions cancel out.

We could use Variance in Turning Angle to illustrate the predictability with which an individual turns, but variance is also calculated based on the mean which does not have much significance for exploration. Large numbers mean that turning angle is very different from the mean, small numbers mean the turning angle is similar to the mean.

Alternatively, we could pick different variables. For one, we could say Duration in the Center is representative of exploration and Latency to enter the Center is representative of boldness. Or we could say Frequency of Zone Alterations is a measure of exploration. Although it could seem like another measure of activity, it is not very correlated with Mean Mobility (0.24). This variable does look like it varies by arena size, so we would need to account for that in some way - maybe include an offset for average body size at each age, or arena size.


```{r exploration alternatives}
library(ggplot2)
ggplot(etho[which(etho$Treatment == "0"),], aes(x = FishName, y = VarTurnAngle)) +
  #facet_wrap(~Treatment) +
  geom_boxplot() +
  theme_classic()
ggplot(etho[which(etho$Treatment == "0"),], aes(x = FishName, y = FreqZAlter)) +
  #facet_wrap(~Treatment) +
  geom_boxplot() +
  theme_classic()
#one outlier individual in Control that crossed zones a lot
ggplot(etho[which(etho$Treatment == "0"),], aes(x = FishName, y = LatencyZ2360)) +
  #facet_wrap(~Treatment) +
  geom_boxplot() +
  theme_classic()
#the ceiling values could make it hard to fit these data

###### Code to check model assumptions
# library(DHARMa)
# library(lme4)
# simulationOutput <- simulateResiduals(fittedModel = glmer() ~ Treatment + scale(Age, center = T, scale = T) + (1|FishName) + (1|Clutch), family=binomial, data=etho), n=250) #250 simulations, but if want higher precision change n>1000; Log transform because trials does not fit a poisson distribution.
# plot(simulationOutput$scaledResiduals) #Expect a flat distribution of the overall residuals, and uniformity in y direction if plotted against any predictor. Looks randomly scattered
# testDispersion(simulationOutput) #if under- or over-dispersed, then p-value<0.05, but then check the dispersion parameter and try to determine what in the model could be the cause and address it there, also check for zero inflation.
# #p=
# testZeroInflation(simulationOutput) #compare expected vs observed zeros, not zero-inflated if p>0.05.
# #p=
# testUniformity(simulationOutput) #check for heteroscedasticity ("a systematic dependency of the dispersion / variance on another variable in the model" Hartig, https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html), which is indicated if dots aren't on the red line and p<0.05.
# #p=
# plot(simulationOutput)
```





